Weakly Nonlinear Analysis of the Standard and Helical Magnetorotational Instabilities in a Taylor Couette Flow

### 2016-08-01 paper/widegap_mri_paper.tex
coefficients plot added

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_paramspace_crit_params_Pm_1E-3}
\caption{Growth rates in the ($\reym, k_z$) plane. Color map shows growth rate found by solving the linear eigenvalue problem for each ($\reym, k_z$) in the grid. The eigenvalue problem was solved for the widegap parameters listed in Table \ref{table:parameters}. Overlaid contours show growth rates at [-2E-3, -1E-3, -1E-5, 1E-5, 1E-3, 2E-3], where dashed contours represent negative values. The gray dotted line shows the interpolated marginal stability curve. The critical parameters $\reym_c = 0.84$ and $k_c = 0.0127$ correspond to the smallest parameter values that yield a zero growth rate.}
\end{figure}

@@

\beq
N^{(\Psi)} = - J(\Psi, \frac{1}{r^2} ( \nabla^2 \Psi - \frac{2}{r} \partial_r\Psi) ) + \frac{2}{\beta} J(A, \frac{1}{r^2} ( \nabla^2 A - \frac{2}{r} \partial_rA) ) - \frac{2}{\beta} \frac{2}{r}B_\phi \partial_z B_\phi  + \frac{2}{r} u_\phi \partial_z u_\phi 
\eeq
\beq
N^{(u)} = \frac{2}{\beta} \frac{1}{r} J(A, B_\phi) - \frac{1}{r} J(\Psi, \uphi) + \frac{2}{\beta}\frac{1}{r^2} B_\phi \partial_z A - \frac{1}{r^2} \uphi \partial_z \Psi 
\eeq
\beq
N^{(A)} = \frac{1}{r} J(A, \psi)
\eeq
\beq
N^{(B)} = \frac{1}{r} J(A, \uphi) + \frac{1}{r} J(B_\phi, \psi) + \frac{1}{r^2} B_\phi \partial_z \psi - \frac{1}{r^2} \uphi \partial_z A 
\label{eq:nonlinear_B}
\eeq

\end{widetext}

@@

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_3D_vel_screenshot.png}
\caption{Visualization of the velocity perturbation field in the fiducial SMRI case.}
\end{figure}

@@

We solve the equations at each order using Dedalus, an open source pseudospectral code (Burns et al. in prep). Because our domain is bounded only radially and not vertically, we solve the radial dimension on a basis of Chebyshev polynomials, and the vertical dimension on a Fourier series basis. We solve Equation \ref{eq:ordere} as a linear eigenvalue problem, and Equations \ref{eq:ordere2} and \ref{eq:ordere3} as linear boundary value problems. %should i mention spurious eigenvals?

@@

a Ginzburg-Landau equation (GLE). The GLE governs the weakly nonlinear amplitude behavior in a wide range of physical systems, including the narrow gap MRI (), Rayleigh-B\'enard convection, and hydrodynamic Taylor Couette flow. We emphasize that this is a model equation, valid only near marginality (see ). The dynamics of the GLE are determined by its coefficients, which are in turn determined by the linear eigenfunctions and nonlinear vectors plotted in Figures \ref{fig:linear_eigenfunctions} and \ref{fig:N2_N3}. Equation \ref{eq:gle} contains three coefficients: $b$, which determines the linear growth rate of the system, $d$, a diffusion coefficient, and $c$, the coefficient of the nonlinear term. When all of the coefficients of Equation \ref{eq:gle} are real, this is known as the real GLE, although the amplitude $\alpha$ remains complex. The real GLE is subject to several well-studied instabilities, including the Ekhaus and Zig-Zag instabilities. When the coefficients are complex, we have the complex GLE, a source of even richer phase dynamics than its real counterpart (e.g. \citei{Aranson:2002}).

@@

>>>>

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_paramspace_crit_params_Pm_1E-3}
\caption{Growth rates in the ($\reym, k_z$) plane. Color map shows growth rate found by solving the linear eigenvalue problem for each ($\reym, k_z$) in the grid. The eigenvalue problem was solved for the widegap parameters listed in Table \ref{table:parameters}. Overlaid contours show growth rates at [-2E-3, -1E-3, -1E-5, 1E-5, 1E-3, 2E-3], where dashed contours represent negative values. The gray dotted line shows the interpolated marginal stability curve. The critical parameters $\reym_c = 0.84$ and $k_c = 0.0127$ correspond to the smallest parameter values that yield a zero growth rate.}\label{fig:growth_rates}
\end{figure}

@@

\beq
N^{(\Psi)} = - J(\Psi, \frac{1}{r^2} ( \nabla^2 \Psi - \frac{2}{r} \partial_r\Psi) ) + \frac{2}{\beta} J(A, \frac{1}{r^2} ( \nabla^2 A - \frac{2}{r} \partial_rA) ) - \frac{2}{\beta} \frac{2}{r}B_\phi \partial_z B_\phi  + \frac{2}{r} u_\phi \partial_z u_\phi 
\eeq
\beq
N^{(u)} = \frac{2}{\beta} \frac{1}{r} J(A, B_\phi) - \frac{1}{r} J(\Psi, \uphi) + \frac{2}{\beta}\frac{1}{r^2} B_\phi \partial_z A - \frac{1}{r^2} \uphi \partial_z \Psi 
\eeq
\beq
N^{(A)} = \frac{1}{r} J(A, \psi)
\eeq
\beq
N^{(B)} = \frac{1}{r} J(A, \uphi) + \frac{1}{r} J(B_\phi, \psi) + \frac{1}{r^2} B_\phi \partial_z \psi - \frac{1}{r^2} \uphi \partial_z A 
\label{eq:nonlinear_B}
\eeq
\end{widetext}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_coeffs_by_Pm.pdf}
\caption{Coefficients.}\label{fig:coefficients}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/widegap_3D_vel_screenshot.png}
\caption{Visualization of the velocity perturbation field in the fiducial SMRI case. The three dimensional path of twenty randomly placed tracer particles is plotted, where the particle velocity is defined as the total (first + second order) velocity perturbation. This is meant to be a visualization only of the velocity field, so the particle motion feels no contribution from the magnetic field.}
\end{figure}

@@

We solve the equations at each order using Dedalus, an open source pseudospectral code (Burns et al. in prep). Because our domain is bounded only radially and not vertically, we solve the radial dimension on a basis of Chebyshev polynomials, and the vertical dimension on a Fourier series basis. We solve Equation \ref{eq:ordere} as a linear eigenvalue problem, and Equations \ref{eq:ordere2} and \ref{eq:ordere3} as linear boundary value problems. %should i mention spurious eigenvals? 
To analyze the MRI system at marginality, we fix the parameters listed in Table \ref{table:parameters} and determine the critical $\reym$ and $k_z$ by repeatedly solving Equation \ref{eq:ordere} to determine the smallest parameter values for which the fastest growing mode is zero. That is, we solve the linear eigenvalue problem for eigenvalues $\sigma = \gamma + i \omega$. Figure \ref{fig:growth_rates} shows linear MRI growth rates $\gamma$ in the $(\reym, k_z)$ plane. For the fiducial standard MRI parameters in Table \ref{table:parameters} we find critical parameters $\reym_c = 0.84$, $k_c = 0.0127$.

@@

a Ginzburg-Landau equation (GLE). The GLE governs the weakly nonlinear amplitude behavior in a wide range of physical systems, including the narrow gap MRI (\citei{Umurhan:2007hs}), Rayleigh-B\'enard convection, and hydrodynamic Taylor Couette flow. We emphasize that this is a model equation, valid only near marginality (see \citei{Cross:1993el}). The dynamics of the GLE are determined by its coefficients, which are in turn determined by the linear eigenfunctions and nonlinear vectors plotted in Figures \ref{fig:linear_eigenfunctions} and \ref{fig:N2_N3}. Equation \ref{eq:gle} contains three coefficients: $b$, which determines the linear growth rate of the system, $d$, a diffusion coefficient, and $c$, the coefficient of the nonlinear term. When all of the coefficients of Equation \ref{eq:gle} are real, this is known as the real GLE, although the amplitude $\alpha$ remains complex. The real GLE is subject to several well-studied instabilities, including the Ekhaus and Zig-Zag instabilities. When the coefficients are complex, we have the complex GLE, a source of even richer phase dynamics than its real counterpart (e.g. \citei{Aranson:2002}).

@@

\bibitem[{Cross \& Hohenberg(1993)}]{Cross:1993el}
Cross, M. C., \& Hohenberg, P. C., 1993, Rev. Mod. Phys. 65, 851
